---
title: "Life Expectancy for Political Candidates"
author: "Ivy Spratt"
format: html
execute: 
  echo: false
---

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(primer.data)
library(tidymodels)
library(broom)
library(knitr)
library(ggplot2)
library(marginaleffects)

# Fit the linear model and tidy results
fit_years <- linear_reg(engine = "lm") |> 
  fit(lived_after ~ sex + election_age + sex*election_age, data = governors) |> 
  tidy(conf.int = TRUE)

# Fit lm model for predictions
fit <- lm(lived_after ~ sex * election_age, data = governors)

# Prepare prediction grid
min_age <- min(governors$election_age, na.rm = TRUE)
max_age <- max(governors$election_age, na.rm = TRUE)

pred_grid <- expand.grid(
  election_age = seq(min_age, max_age, length.out = 100),
  sex = c("Female", "Male")
)

pred_grid <- pred_grid %>%
  mutate(
    pred = predict(fit, newdata = pred_grid, interval = "confidence")[, "fit"],
    conf_low = predict(fit, newdata = pred_grid, interval = "confidence")[, "lwr"],
    conf_high = predict(fit, newdata = pred_grid, interval = "confidence")[, "upr"]
  )

# Plot predicted life expectancy
ggplot(pred_grid, aes(x = election_age, y = pred, color = sex, fill = sex)) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = conf_low, ymax = conf_high), alpha = 0.2, color = NA) +
  labs(
    title = "Predicted Life Expectancy after Election by Age and Sex",
    subtitle = "Male candidates lose more years of life expectancy with increasing election age compared to female candidates.",
    caption = "Data source: Barfort et al. (2020), 'Longevity Returns to Political Office'",
    x = "Age at Election (years)",
    y = "Predicted Years Lived After Election"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 12, color = "gray40"),
    plot.caption = element_text(size = 10, color = "gray60"),
    axis.title = element_text(size = 12)
  )
```

Being a government official is demanding. Using the data from 1000+ governors in the United States, we attempt to create a model predicting life span of future candidates. Since the data only includes the governor position, it may be not as relevant to other positions in government. A 50-year-old male political candidate can expect to live about 30 years after the election, plus-or-minus about 1 year.

## Fitted Model Structure

The model we are fitting can be expressed as:

$$
\text{lived\_after} = \beta_0 + \beta_1 \cdot \text{sex} + \beta_2 \cdot \text{election\_age} + \beta_3 \cdot (\text{sex} \times \text{election\_age}) + \epsilon
$$

```{r}
kable(
  fit_years, 
  caption = "Table 1: Regression Model Results (Data Source: Barfort et al., 2020, 'Longevity Returns to Political Office')"
)
```
